C/O and Snowline Locations in Protoplanetary Disks: The Effect of Radial Drift and Viscous Gas Accretion

### 2015-07-27 tex/draft/ms_kedit_tedit.tex
update

An important consequence of snowline formations in disks is that disks are expected to present different carbon-to-oxygen (C/O) ratios in the gas and in icy dust mantles at different disk radii. This effect was quantified by  \citet{oberg11}, who considered the fact that the main carries of carbon and oxygen, i.e. H$_2$O, CO$_2$ and CO, have different condensation temperatures. This changes the relative abundance of C and O in gaseous and solid form as a function of the snowline location of the volatiles mentioned above. \citet{oberg11} calculated analytically the C/O ratio in gas in dust as a function of semimajor axis for passive protoplanetary disks and found a gas C/O ratio of order unity between the CO$_2$ and CO snowlines, where oxygen gas is highly depleted. This effect have also been observed in some exoplanet atmospheres, where superstellar C/O ratios have been detected (e.g., WASP-12b, \citealt{madhu11}).
%Superstellar C/O ratios have also been detected in the atmospheres of gas giants such as WASP-12b \citep{madhu11}.

@@

\citet{oberg11} assume a static disk with passive chemistry. In reality, additional dynamical and chemical processes affect the snowline locations and the resulting C/O ratio. Several works have addressed some of these effects. \citet{madhu14} use a steady-state active disk model that includes planetary migration and use the C/O ratio to constrain migration mechanisms. \citealt{alidib14} calculate the C/O ratio throughout the disk by incorporating the evolution of solids, i.e. radial drift, sublimation and grain coagulation, as well as the diffusion of volatile vapors. \citet{thiabaud15} consider additional volatile species in their chemical network, such as N$_2$ or NH$_3$, and find that the C/O ratio may be enriched by up to four times the Solar value in certain parts of the disk.

@@

To understand the separate effects of radial drift, radial movement of gas throughout the disk due to gas accretion, and accretion heating, we use four separate disk models:  \textit{passive disk}, which is only irradiated by the central star and does not experience gas accretion or accretional heating; \textit{active disk with passive temperature profile} (hereafter \textit{active disk}), in which the gas is accreting onto the central star causing the gas surface density to decrease with time, but which does not experience accretion heating; \textit{active disk at steady-state}, for which the gas surface density $\Sigma_{\rm d}$ and mass flux $\dot{M}$ are constant in time and independent of semimajor axis \footnote{However, we also include stellar irradiation in the disk thermal structure, which changes the steady-state solution as we explain further in this section.}; and \textit{static disk}, which has a passive temperature profile and does not take into account gas accretion onto the central star or radial drift.

@@

Solid particles in orbit their host star at the Keplerian velocity $v_{\rm k} \equiv \Omega_{\rm k} r$. The gas, however, experiences an additional pressure gradient, which causes it to rotate at sub-Keplerian velocity \citep{weidenschilling77}. Dust grains that are large enough thus experience a headwind, which removes angular momentum, causing the solids to spiral inwards and fall onto the host star. Small particles are well-coupled to the gas, while large planetesimals are decoupled from the gas. From \citet{chiang10}, the extent of coupling is quantified by the dimensionless stopping time, $\tau_{\rm s} \equiv \Omega_{\rm k} t_{\rm s}$, where $t_{\rm s}$ is
\begin{equation}
\label{eq:ts}
t_{\rm s}= \left\{
\begin{array}{l l}
\rho_{\rm s} s / (\rho_{\rm d} c_{\rm d}), & \quad s < 9 \lambda/4 \,\,\,\ \text{Epstein drag} \\
4 \rho_{\rm s} s^2 / (9 \rho_{\rm d} c_{\rm d} \lambda), & \quad s < 9 \lambda/4, \,\text{Re} \lesssim 1 \,\,\,\ \text{Stokes drag.}
\end{array} 
\right.
\end{equation}
Here $\rho_{\rm d}$ is the gas midplane density, $\rho_{\rm s}=2$ g cm$^{-3}$ is the density of a solid particle, $s$ is the particle size, $\lambda$ is the mean free path, and Re is the Reynolds number.

For a passive disk, the radial drift velocity can be approximated analytically as
\begin{equation}
\label{eq:rdotpas}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big),
\end{equation}
where
\begin{equation}
\label{eq:eta}
\eta \equiv - \frac{\partial P_{\rm d}/\partial \ln r}{2 \rho_{\rm d} v_{\rm k}^2} \approx \frac{c_{\rm d}^2}{2 v_k^2}
\end{equation}
and $P_{\rm d} = \rho_{\rm d} c_{\rm d}^2$ is the disk midplane pressure.

@@

\textit{Radial drift timescale.} To order of magnitude, the radial drift timescale can be estimated as 
\begin{equation}
\label{eq:tdrift}
t_{\rm r, drift} \sim \Big|\frac{r}{\dot{r}}\Big|,
\end{equation}
where $\dot{r}$ is the radial drift velocity given by Equation (\ref{eq:rdotpas}) for a passive disk and by Equation (\ref{eq:rdotact}) for an active disk.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{drift_timescales_betaS1_gas_acc_new.pdf}
%\vspace{-0.5in}
\caption{Relevant timescales for dynamical effects in the desorption process: $t_{\rm r, drift}$ (solid lines), $t_{\rm des}$ (dashed lines) and $t_{\rm gas, acc}$ (dotted lines). The timescales are calculated at three representative locations, i.e. the H$_2$O, CO$_2$ and CO snowlines in the static disk. For our choice of parameters, the snowlines are located at $\sim$0.7 AU (blue lines), $\sim$8.6 AU (green lines) and $\sim$59 AU (red lines), respectively. The horizontal dashed line represents a typical disk lifetime of 3 Myr. Radial drift and gas accretion affect desorption in the regions where their respective timescales, i.e. $t_{\rm r, drift}$ and $t_{\rm gas, acc}$, are comparable to the desorption timescale $t_{\rm des}$.} 
\label{fig:timescales}
\end{figure}

For simplicity purposes, we calculate the radial drift timescale, $t_{\rm r, drift}$, for a passive disk in this section, but most of our conclusions hold true for an active disk as well. Figure \ref{fig:timescales} shows $t_{\rm des}$, $t_{\rm r, drift}$ and $t_{\rm gas, acc}$ as a function of particle size at three different locations in the disk, corresponding to the H$_2$O, CO$_2$ and CO snowlines in the static disk. As expected, micron-sized particles desorb on very short timescales of $\sim 1-1000$ years in the close vicinity of their respective snowlines, since the desorption rate depends exponentially on temperature and hence on disk location (see Equation \ref{eq:Rdes}).  On the other hand, their radial drift timescale exceeds the typical disk lifetime of a few Myr by several orders of magnitude due to their strong coupling with the gas. Thus for small particles in a passive disk, the snowline locations and the C/O ratio are the same as for a static disk (see Figure 1 from \citealt{oberg11}) \footnote{This is not true for an active disk, however, where gas accretion causes even micron-sized particles to drift significantly before desorbing, as we show in section \ref{sec:snowlines}.}. At the other extreme, kilometer-sized particles are unaffected by gas drag and have long desorption timescales ($\gg$1 Myr ), and the snowline locations and C/O ratio remain unchanged in this case as well. This is true for both passive and active disks, since large planetesimals are decoupled from the gas and hence unaffected by gas accretion onto the host star.

Of particular interest for our purposes is the particle size regime for which (1) $t_{\rm r, drift} \lesssim t_{\rm des} \lesssim t_{\rm d}$ ($t_{\rm d}=3$ Myr is the disk lifetime), i.e. for $\sim$$0.5$ cm $\lesssim s \lesssim$ $1000$ cm, or (2) $t_{\rm gas, acc} \lesssim t_{\rm des} \lesssim t_{\rm d}$, i.e. for $\sim$$0.1$ cm $\lesssim s \lesssim$ $10$ cm. In these cases, radial drift or gas accretion (or both) are faster than thermal desorption. We note that $t_{\rm gas, acc}<t_{\rm d}$ always holds true. Particles of sizes that satisfy the requirements above will drift significantly due to radial drift or gas accretion before desorbing, thus moving the H$_2$O, CO$_2$ and CO snowlines closer towards the central star and changing the C/O ratio throughout the disk. We quantify these effects in sections \ref{sec:snowlines} and \ref{sec:COratio}.

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=0.7\textwidth]{desorption_distance_passive_active_colorbar_test.pdf}
%\vspace{-0.5in}
\caption{Desorption distance (if a grain fully desorbs) or final distance (if a grain does not fully desorb), as a function of a particle's initial location in the disk, for a range of particle sizes, and for both a passive disk (left panels) and an active disk (right panels). The desorption distance is calculated for particles composed of H$_2$O (top panels), CO$_2$ (middle panels) and CO (bottom panels). The particle size increases from $10^{-3}$ cm to $10^8$ cm as indicated by the color bar. For a particle of a given initial size that entirely desorbs during $t_{\rm d}=3$ Myr, the desorption distance is the same regardless of the particle's initial location.} 
\label{fig:snowlines}
\end{figure*}

@@

\begin{subeqnarray}
\label{eq:ddt}
\frac{ds}{dt} &= & - \frac{3 \mu_x m_{\rm p}}{\rho_{\rm s}} N_x R_{\rm des, x}  \slabel{eq:dsdt} \\
\frac{dr}{dt} &=& \dot{r} \slabel{eq:drdt},
\end{subeqnarray}
where the desorption rate $R_{\rm des, x}$ for each particle type (i.e., composed of H$_2$O, CO$_2$ or CO) is evaluated at $T=T_{\rm d}(r)$, and the radial drift velocity $\dot{r}$ is given by Equation (\ref{eq:rdotpas}) for a passive disk and Equation (\ref{eq:rdotact}) for an active disk. Equations (\ref{eq:dsdt}) and (\ref{eq:drdt}) describe the coupled desorption and radial drift, and can be derived straightforwardly from Equation (\ref{eq:tdes}). Our boundary conditions are $s(t_0)=s_0$, $r(t_0)=r_0$, and $s(t_{\rm d})=0$, where $t_0$ is the initial time at which we start the integration and $r_0$ is the initial location of the particle. We choose $t_0=1$ year, but our result is independent on the initial integration time as long as $t_0 \ll t_{\rm d}$.

@@

We define the final position of a grain as the disk location it has reached after $t_{\rm d}=3$ Myr, or the radius at which it completely desorbs if that happens after a time shorter than 3 Myr.  Figure \ref{fig:snowlines} shows our results for H$_2$O, CO$_2$ and CO particles, for both a passive and an active disk. The results for the steady-state active disk are qualitatively similar to the active disk with a passive disk temperature gradient and therefore not shown. %We do not show the equivalent result for the steady-state active disk since The trends are qualitatively the same as for the active disk with the temperature profile given by Equation (\ref{eq:diskT}), but we discuss the C/O ratio steady-state active disk results in Section \ref{sec:COratio}. 
Kilometer-sized bodies do not drift or desorb during the disk lifetime neither for a passive nor for an active disk. Similarly, micron- to mm-sized particles in the passive disk do not drift and only %or 
desorb if %unless 
they are located inside the static snowlines. %This is not in contradiction with the desorption timescales for small particles from Figure \ref{fig:timescales}, since small grains only desorb fast at or nearby their respective snowlines, as mentioned in Section \ref{sec:timescales}.  [don't think this is needed, but Til or Ruth might think otherwise]
In an active disk, however, micron-to mm-sized grains do drift significantly since they move at the same velocity as the accreting gas. For $0.5$ cm $\lesssim s_0 \lesssim$ 700 cm in a passive disk and $0.001$ cm $\lesssim s_0 \lesssim$ 700 cm in an active disk, we notice that particles of initial size $s_0$ desorb at a %fixed distance 
particle size specific radius $r_{\rm des}$ regardless of their original location in the disk. In fact, the only grains that will both drift and evaporate are those that reach their fixed final location (represented by the horizontal curves in Figure \ref{fig:snowlines}) within the disk lifetime. We show in section \ref{sec:COratio} that this result is essential in determining the C/O ratio throughout the disk for different particle sizes.

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{s_t_a.pdf}
%\vspace{-0.5in}
\caption{Left panels: size of desorbing H$_2$O particles as a function of time, for different initial particle sizes and for three initial locations in a passive disk: 1 AU (top left), 10 AU (middle left) and 100 AU (bottom left). Particles desorb almost instantaneously. Right panel: fractional mass of the desorbing particles as a function of the particle's location as it drifts, for different initial particle sizes, and at the same initial locations presented in the left panel. Particles lose most of their mass very close to the distance at which they fully desorb.}
\label{fig:s_t_a}
\end{figure*}

Intuitively, this fixed $r_{\rm des}$ should be the location in the disk for which $t_{\rm r, drift} \sim t_{\rm des}$, given an initial particle size. We can calculate this location analytically by equating Equations (\ref{eq:tdes}) and (\ref{eq:tdrift}) and solving for $r=r_{\rm des}(s)$ for a given particle size $s$. Figure \ref{fig:an_vs_actual} shows $r_{\rm des}$ calculated analytically using the prescription above as a function of the actual desorption distance calculated numerically. We display this result for the range of particle sizes that desorb at a fixed distance in a passive and an active disk (see Figure \ref{fig:snowlines}). We notice that the analytic approximation accurately reproduces the numerical result for most cases of interest, but it slightly deviates for particles larger than $s \gtrsim10$ cm. For small particles with $\tau_{\rm s} \ll 1$, $t_{\rm r, drift}$ is a power-law in $r$ (for our parameters, $t_{\rm r, drift} \propto r^{-1/14}$ for the passive disk), and the Equation set (\ref{eq:ddt}) has an explicit analytic solution (see \App{app:tdriftan}). Once particles are large enough so that $\tau_{\rm s} \sim 1$, $t_{\rm r, drift}$ has a more complicated dependence on $r$ (see Equation \ref{eq:ts}), and the coupled drift-desorption differential equations have to be integrated numerically to obtain an accurate result.

@@

Figure \ref{fig:s_t_a}, left panels, shows the size evolution with time for H$_2$O particles of various initial sizes, starting at three different initial locations in a passive disk \footnote{Our conclusions remain valid for an active disk and for particles composed of CO$_2$ or CO.}.  Once solid H$_2$O particles begin to evaporate, they do so almost instantly for all explored particle sizes and initial locations. %, although the time $t_{\rm des}$ at which a particle of a given initial size desorbs depends on its initial distance. 
%A particle located at the initial time $t_0$ at a distance such that $t_{\rm des}>t_{\rm d}$ will therefore not desorb during the disk lifetime, but our model assumes that particles drift continuously at any location in the disk. [not sure you need to explain this]
The right panels of Figure \ref{fig:s_t_a} show that the drifting grains lose most of their mass in a very narrow distance range; moreover, this distance is the same for a given initial particle size, no matter where the particle started drifting at the time $t_0$ when the simulation is started. Figure \ref{fig:s_t_a} thus demonstrates that solid particles that drift and fully desorb during the lifetime of the protoplanetary disk do so (1) instantaneously, and (2) at a fixed stellocentric distance, regardless of their initial location in the disk. 
It follows that the H$_2$O, CO$_2$ and CO snowlines are fixed for a given initial particle size and disk model (passive or active). The C/O ratio will then only depend on disk properties, grain size, and the abundance of H$_2$O, CO$_2$ and CO relative to the H$_2$ abundance in the disk midplane, and {\it not} on the disk age when only considering drift, accretion and desorption.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=\textwidth]{time_plots.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for particles of initial size $s_0=1$ m (left panels) and $s_0=10$ cm (right panels), for grains composed of H$_2$O (top panels), CO$_2$ (middle panels) and CO (bottom panels). The evolution is shown at four representative timescales: $10^4$ yr (cyan curve), $10^5$ yr (red curve), 1 Myr (green curve), and 3 Myr, the disk lifetime (blue curve). For a given particle size, the desorption distance, and hence the H$_2$O, CO$_2$ and CO snowlines, have the same location regardless of the time at which the simulation is stopped.}
\label{fig:timeplots}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{desorption_distance_varying_Md.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for H$_2$O particles of initial size of 1 m, for total disk masses $M_{\rm d}=0.1 M_{\odot}$ (solid lines) and $M_{\rm d}=0.01 M_{\odot}$ (dashed lines). The timescales of the simulations and their color code are the same as in Figure \ref{fig:timeplots}. A lower disk mass does not change the snowline location.}
\label{fig:varMd}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{desorption_distance_transition_disk_1000.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for H$_2$O particles of initial size of 1 m, for our fiducial disk (solid lines) and for a transition disk with an inner cavity at $r_0=4$ AU (dashed lines). The timescales of the simulations and their color code are the same as in Figure \ref{fig:timeplots}. Particles that start inside the cavity drift towards the original snowline, while particles that start outside the gap stop shortly after crossing the gap edge.}
\label{fig:cavity}
\end{figure}

@@

We summarize in Table 1 the potential physical and chemical processes occurring in disks and their effect on snowline locations. For the sake of completion, Table 1 also includes the processes addressed in this paper, i.e. radial drift and gas accretion. The neglected effects are discussed in more detail below.

>>>>

An important consequence of snowline formations in disks is that disks are expected to present different carbon-to-oxygen (C/O) ratios in the gas and in icy dust mantles at different disk radii. This effect was quantified by  \citet{oberg11}, who considered the fact that the main carries of carbon and oxygen, i.e. H$_2$O, CO$_2$ and CO, have different condensation temperatures. This changes the relative abundance of C and O in gaseous and solid form as a function of the snowline location of the volatiles mentioned above. \citet{oberg11} calculated analytically the C/O ratio in gas in dust as a function of semimajor axis for passive protoplanetary disks and found a gas C/O ratio of order unity between the CO$_2$ and CO snowlines, where oxygen gas is highly depleted. This effect has also been observed in some exoplanet atmospheres, where superstellar C/O ratios have been detected (e.g., WASP-12b, \citealt{madhu11}).
%Superstellar C/O ratios have also been detected in the atmospheres of gas giants such as WASP-12b \citep{madhu11}.

@@

\citet{oberg11} assumed a static disk with passive chemistry. In reality, additional dynamical and chemical processes affect the snowline locations and the resulting C/O ratio. Several works have addressed some of these effects. \citet{madhu14} use a steady-state active disk model that includes planetary migration and use the C/O ratio to constrain migration mechanisms. \citealt{alidib14} calculate the C/O ratio throughout the disk by incorporating the evolution of solids, i.e. radial drift, sublimation and grain coagulation, as well as the diffusion of volatile vapors. \citet{thiabaud15} consider additional volatile species in their chemical network, such as N$_2$ or NH$_3$, and find that the C/O ratio may be enriched by up to four times the Solar value in certain parts of the disk.

@@

To understand the separate effects of radial drift, radial movement of gas throughout the disk due to gas accretion, and accretion heating, we use four separate disk models:  \textit{passive disk}, which is only irradiated by the central star and does not experience gas accretion or accretional heating; \textit{active disk with passive temperature profile} (hereafter \textit{active disk}), in which the gas is accreting onto the central star causing the gas surface density to decrease with time, but which does not experience accretion heating; \textit{active disk at steady-state}, for which the gas surface density $\Sigma_{\rm d}$ and mass flux $\dot{M}$ are constant in time and independent of semimajor axis\footnote{However, we also include stellar irradiation in the disk thermal structure, which changes the steady-state solution as we explain further in this section.}; and \textit{static disk}, which has a passive temperature profile and does not take into account gas accretion onto the central star or radial drift.

@@

Solid particles orbit their host star at the Keplerian velocity $v_{\rm k} \equiv \Omega_{\rm k} r$. The gas, however, experiences an additional pressure gradient, which causes it to rotate at sub-Keplerian velocity \citep{weidenschilling77}. Dust grains that are large enough thus experience a headwind, which removes angular momentum, causing the solids to spiral inwards and fall onto the host star. Small particles are well-coupled to the gas, while large planetesimals are decoupled from the gas. From \citet{chiang10}, the extent of coupling is quantified by the dimensionless stopping time, $\tau_{\rm s} \equiv \Omega_{\rm k} t_{\rm s}$, where $t_{\rm s}$ is
\begin{equation}
\label{eq:ts}
t_{\rm s}= \left\{
\begin{array}{l l}
\rho_{\rm s} s / (\rho_{\rm d} c_{\rm d}), & \quad s < 9 \lambda/4 \,\,\,\ \text{Epstein drag} \\
4 \rho_{\rm s} s^2 / (9 \rho_{\rm d} c_{\rm d} \lambda), & \quad s < 9 \lambda/4, \,\text{Re} \lesssim 1 \,\,\,\ \text{Stokes drag.}
\end{array} 
\right.
\end{equation}
Here $\rho_{\rm d}$ is the gas midplane density, $\rho_{\rm s}=2$ g cm$^{-3}$ is the density of a solid particle, $s$ is the particle size, $\lambda$ is the mean free path, and Re is the Reynolds number.

For a passive disk, the radial drift velocity can be approximated as
\begin{equation}
\label{eq:rdotpas}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big),
\end{equation}
where
\begin{equation}
\label{eq:eta}
\eta \equiv - \frac{\partial P_{\rm d}/\partial \ln r}{2 \rho_{\rm d} v_{\rm k}^2} \approx \frac{c_{\rm d}^2}{2 v_k^2}
\end{equation}
and $P_{\rm d} = \rho_{\rm d} c_{\rm d}^2$ is the disk midplane pressure.

@@

\textit{Radial drift timescale.} To order of magnitude, the radial drift timescale can be estimated as 
\begin{equation}
\label{eq:tdrift}
t_{\rm drift} \sim \Big|\frac{r}{\dot{r}}\Big|,
\end{equation}
where $\dot{r}$ is the radial drift velocity given by Equation (\ref{eq:rdotpas}) for a passive disk and by Equation (\ref{eq:rdotact}) for an active disk.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{drift_timescales_betaS1_gas_acc_new.pdf}
%\vspace{-0.5in}
\caption{Relevant timescales for dynamical effects in the desorption process: $t_{\rm drift}$ (solid lines), $t_{\rm des}$ (dashed lines) and $t_{\rm gas, acc}$ (dotted lines). The timescales are calculated at three representative locations, i.e. the H$_2$O, CO$_2$ and CO snowlines in the static disk. For our choice of parameters, the snowlines are located at $\sim$0.7 AU (blue lines), $\sim$8.6 AU (green lines) and $\sim$59 AU (red lines), respectively. The horizontal dashed line represents a typical disk lifetime of 3 Myr. Radial drift and gas accretion affect desorption in the regions where their respective timescales, i.e. $t_{\rm drift}$ and $t_{\rm gas, acc}$, are comparable to the desorption timescale $t_{\rm des}$.} 
\label{fig:timescales}
\end{figure}

For simplicity, we calculate the radial drift timescale, $t_{\rm drift}$, for a passive disk in this section, but most of our conclusions hold true for an active disk as well. Figure \ref{fig:timescales} shows $t_{\rm des}$, $t_{\rm drift}$ and $t_{\rm gas, acc}$ as a function of particle size at three different locations in the disk, corresponding to the H$_2$O, CO$_2$ and CO snowlines in the static disk. As expected, micron-sized particles desorb on very short timescales of $\sim 1-1000$ years in the close vicinity of their respective snowlines, since the desorption rate depends exponentially on temperature and hence on disk location (see Equation \ref{eq:Rdes}).  On the other hand, their radial drift timescale exceeds the typical disk lifetime of a few Myr by several orders of magnitude due to their strong coupling with the gas. Thus for small particles in a passive disk, the snowline locations and the C/O ratio are the same as for a static disk (see Figure 1 from \citealt{oberg11}) \footnote{This is not true for an active disk, however, where gas accretion causes even micron-sized particles to drift significantly before desorbing, as we show in section \ref{sec:snowlines}.}. At the other extreme, kilometer-sized particles are unaffected by gas drag and have long desorption timescales ($\gg$1 Myr ), and the snowline locations and C/O ratio remain unchanged in this case as well. This is true for both passive and active disks, since large planetesimals are decoupled from the gas and hence unaffected by gas accretion onto the host star.

Of particular interest for our purposes is the particle size regime for which (1) $t_{\rm drift} \lesssim t_{\rm des} \lesssim t_{\rm d}$ ($t_{\rm d}=3$ Myr is the disk lifetime), i.e. for $\sim$$0.5$ cm $\lesssim s \lesssim$ $1000$ cm, or (2) $t_{\rm gas, acc} \lesssim t_{\rm des} \lesssim t_{\rm d}$, i.e. for $\sim$$0.1$ cm $\lesssim s \lesssim$ $10$ cm. In these cases, radial drift or gas accretion (or both) are faster than thermal desorption. We note that $t_{\rm gas, acc}<t_{\rm d}$ always holds true. Particles of sizes that satisfy the requirements above will drift significantly due to radial drift or gas accretion before desorbing, thus moving the H$_2$O, CO$_2$ and CO snowlines closer towards the central star and changing the C/O ratio throughout the disk. We quantify these effects in sections \ref{sec:snowlines} and \ref{sec:COratio}.

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=0.7\textwidth]{desorption_distance_passive_active_colorbar_test2.pdf}
%\vspace{-0.5in}
\caption{Desorption distance (if a grain fully desorbs) or final distance (if a grain does not fully desorb), as a function of a particle's initial location in the disk, for a range of particle sizes, and for both a passive disk (left panels) and an active disk (right panels). The desorption distance is calculated for particles composed of H$_2$O (top panels), CO$_2$ (middle panels) and CO (bottom panels). The particle size increases from $10^{-3}$ cm to $10^8$ cm as indicated by the color bar. For a particle of a given initial size that entirely desorbs during $t_{\rm d}=3$ Myr, the desorption distance is the same regardless of the particle's initial location.} 
\label{fig:snowlines}
\end{figure*}

@@

\begin{subeqnarray}
\label{eq:ddt}
\frac{ds}{dt} &= & - \frac{3 \mu_x m_{\rm p}}{\rho_{\rm s}} N_x R_{\rm des, x}  \slabel{eq:dsdt} \\
\frac{dr}{dt} &=& \dot{r} \slabel{eq:drdt},
\end{subeqnarray}
where the desorption rate $R_{\rm des, x}$ for each particle type (i.e., composed of H$_2$O, CO$_2$ or CO) is evaluated at $T=T_{\rm d}(r)$, and the radial drift velocity $\dot{r}$ is given by Equation (\ref{eq:rdotpas}) for a passive disk and Equation (\ref{eq:rdotact}) for an active disk. Equations (\ref{eq:dsdt}) and (\ref{eq:drdt}) describe the coupled desorption and radial drift, and can be derived straightforwardly from Equation (\ref{eq:tdes}). Our initial conditions are $s(t_0)=s_0$ and $r(t_0)=r_0$, where $t_0$ is the initial time at which we start the integration and $r_0$ is the initial location of the particle.  We choose $t_0=1$ year, but our result is independent on the initial integration time as long as $t_0 \ll t_{\rm d}$. The desorption timescale $t_{\rm des}$ will then satisfy $s(t_{\rm des})=0$, from which we can determine the desorption distance $r_{\rm des}=r(t_{\rm des})$.

@@

We define the final position of a grain as the disk location it has reached after $t_{\rm d}=3$ Myr, or the radius at which it completely desorbs if that happens after a time shorter than 3 Myr.  Figure \ref{fig:snowlines} shows our results for H$_2$O, CO$_2$ and CO particles, for both a passive and an active disk. The results for the steady-state active disk are qualitatively similar to the active disk with a passive disk temperature gradient and therefore not shown. %We do not show the equivalent result for the steady-state active disk since The trends are qualitatively the same as for the active disk with the temperature profile given by Equation (\ref{eq:diskT}), but we discuss the C/O ratio steady-state active disk results in Section \ref{sec:COratio}. 
Kilometer-sized bodies do not drift or desorb during the disk lifetime neither for a passive nor for an active disk. Similarly, micron- to mm-sized particles in the passive disk do not drift and only %or 
desorb if %unless 
they are located inside the static snowlines. %This is not in contradiction with the desorption timescales for small particles from Figure \ref{fig:timescales}, since small grains only desorb fast at or nearby their respective snowlines, as mentioned in Section \ref{sec:timescales}.  [don't think this is needed, but Til or Ruth might think otherwise]
In an active disk, however, micron-to mm-sized grains do drift significantly since they move at the same velocity as the accreting gas. For $0.5$ cm $\lesssim s_0 \lesssim$ 700 cm in a passive disk and $0.001$ cm $\lesssim s_0 \lesssim$ 700 cm in an active disk, we notice that particles of initial size $s_0$ desorb at a %fixed distance 
particle size dependent radius $r_{\rm des}$ regardless of their original location in the disk. In fact, the only grains that will both drift and evaporate are those that reach their fixed final location (represented by the horizontal curves in Figure \ref{fig:snowlines}) within the disk lifetime. We show in section \ref{sec:COratio} that this result is essential in determining the C/O ratio throughout the disk for different particle sizes.

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{s_t_a.pdf}
%\vspace{-0.5in}
\caption{Left panels: size of desorbing H$_2$O particles as a function of time, for different initial particle sizes and for three initial locations in a passive disk: 1 AU (top left), 10 AU (middle left) and 100 AU (bottom left). Particles desorb almost instantaneously. Right panel: fractional mass of the desorbing particles as a function of the particle's location as it drifts, for different initial particle sizes, and at the same initial locations presented in the left panel. Particles lose most of their mass very close to the distance at which they fully desorb. The static H$_2$O snowline is shown for reference (dashed vertical lines).}
\label{fig:s_t_a}
\end{figure*}

Intuitively, this fixed $r_{\rm des}$ should be the location in the disk for which $t_{\rm drift} \sim t_{\rm des}$, given an initial particle size. We can calculate this location analytically by equating Equations (\ref{eq:tdes}) and (\ref{eq:tdrift}) and solving for $r=r_{\rm des}(s)$ for a given particle size $s$. Figure \ref{fig:an_vs_actual} shows $r_{\rm des}$ calculated analytically using the prescription above as a function of the actual desorption distance calculated numerically. We display this result for the range of particle sizes that desorb at a fixed distance in a passive and an active disk (see Figure \ref{fig:snowlines}). We notice that the analytic approximation accurately reproduces the numerical result for most cases of interest, but it slightly deviates for particles larger than $s \gtrsim10$ cm. For small particles with $\tau_{\rm s} \ll 1$, $t_{\rm drift}$ is a power-law in $r$ (for our parameters, $t_{\rm drift} \propto r^{-1/14}$ for the passive disk), and the Equation set (\ref{eq:ddt}) has an explicit analytic solution (see \App{app:tdriftan}). Once particles are large enough so that $\tau_{\rm s} \sim 1$, $t_{\rm drift}$ has a more complicated dependence on $r$ (see Equation \ref{eq:ts}), and the coupled drift-desorption differential equations have to be integrated numerically to obtain an accurate result.

@@

Figure \ref{fig:s_t_a}, left panels, shows the size evolution with time for H$_2$O particles of various initial sizes, starting at three different initial locations in a passive disk\footnote{Our conclusions remain valid for an active disk and for particles composed of CO$_2$ or CO.}.  Once solid H$_2$O particles begin to evaporate, they do so almost instantly for all explored particle sizes and initial locations. %, although the time $t_{\rm des}$ at which a particle of a given initial size desorbs depends on its initial distance. 
%A particle located at the initial time $t_0$ at a distance such that $t_{\rm des}>t_{\rm d}$ will therefore not desorb during the disk lifetime, but our model assumes that particles drift continuously at any location in the disk. [not sure you need to explain this]
The right panels of Figure \ref{fig:s_t_a} show that the drifting grains lose most of their mass in a very narrow distance range; moreover, this distance is the same for a given initial particle size, no matter where the particle started drifting at the time $t_0$ when the simulation is started. Figure \ref{fig:s_t_a} thus demonstrates that solid particles that drift and fully desorb during the lifetime of the protoplanetary disk do so (1) instantaneously, and (2) at a fixed stellocentric distance, regardless of their initial location in the disk. 
It follows that the H$_2$O, CO$_2$ and CO snowlines are fixed for a given initial particle size and disk model (passive or active). The C/O ratio will then only depend on disk properties, grain size, and the abundance of H$_2$O, CO$_2$ and CO relative to the H$_2$ abundance in the disk midplane, and {\it not} directly on the disk age when only considering drift, accretion and desorption.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=\textwidth]{time_plots.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for particles of initial size $s_0=1$ m (left panels) and $s_0=10$ cm (right panels), for grains composed of H$_2$O (top panels), CO$_2$ (middle panels) and CO (bottom panels). The evolution is shown at four representative timescales: $10^4$ yr (cyan curve), $10^5$ yr (purple curve), 1 Myr (blue curve), and 3 Myr, the disk lifetime (orange curve). For a given particle size, the desorption distance, and hence the H$_2$O, CO$_2$ and CO snowlines, have the same location regardless of the time at which the simulation is stopped.}
\label{fig:timeplots}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{desorption_distance_varying_Md.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for H$_2$O particles of initial size of 1 m, for total disk masses $M_{\rm d}=0.1 M_{\odot}$ (solid lines) and $M_{\rm d}=0.01 M_{\odot}$ (dashed lines). The timescales at which we stop the simulations are $10^4$ yr (cyan curve), $10^5$ yr (red curve), 1 Myr (green curve) and 3 Myr (blue curve). A lower disk mass does not change the snowline location.}
\label{fig:varMd}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{desorption_distance_transition_disk_1000.pdf}
%\vspace{-0.5in}
\caption{Desorption / final distance as a function of initial position in the disk for H$_2$O particles of initial size of 1 m, for our fiducial disk (solid lines) and for a transition disk with an inner cavity at $r_0=4$ AU (dashed lines). The timescales of the simulations and their color code are the same as in Figure \ref{fig:varMd}. Particles that start inside the cavity drift towards the original snowline, while particles that start outside the gap stop shortly after crossing the gap edge.}
\label{fig:cavity}
\end{figure}

@@

We summarize in Table 1 the potential physical and chemical processes occurring in disks and their effect on snowline locations compared to a static disk. For the sake of completeness, Table 1 also includes the processes addressed in this paper, i.e. radial drift and gas accretion. The neglected effects are discussed in more detail below.
